---
title: "Reproducible Research Course Project 2"
author: "Sander Schrieken"
date: "10/30/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events.

## Questions

1. Does the analysis address the question of which types of events are most harmful to population health?
2. Does the analysis address the question of which types of events have the greatest economic consequences?


## Importing Data

Does the analysis start from the raw data file (i.e. the original .csv.bz2 file)?

The code below downloads the data from the Coursera webpage for the Reproducible Research Course Project (#2). This is the [link](https://www.coursera.org/learn/reproducible-research/peer/OMZ37/course-project-2) to that page.

```{r download_(meta)_data, message=FALSE, warning=FALSE, eval=FALSE, cache=TRUE}
Download_Date <- Sys.Date() # First, fix the date we download the data!

source_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(source_url, destfile = "StormData.csv.bz2", 
              method = "curl")

source_url_2 <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
download.file(source_url_2, destfile = "repdata-peer2_doc-pd01016005curr.pdf", 
              method = "curl")

source_url_3 <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf"
download.file(source_url_3,
              destfile = "repdata-peer2_doc-NCDC Storm Events-FAQ Page.pdf",
              method = "curl")
```

```{r rename_meta_data_files, echo=TRUE, results="hide", message=FALSE, warning=FALSE, cache=TRUE}
# The names of the files are unwieldy; we should consider renaming them...
file.rename("repdata-peer2_doc-pd01016005curr.pdf", "Storm Data Documentation.pdf")
file.rename("repdata-peer2_doc-NCDC Storm Events-FAQ Page.pdf", "Storm Events FAQs.pdf")
```


## Data Processing

Is there a section titled "Data Processing" that describes how the data were loaded into R and processed for analysis?

Reading the downloaded data into R can be done with `read.csv()`. By using this function we don't have to unzip the compressed file separately, since `read.csv()` can handle this for us. The following code reads the downloaded data into R. We will _NOT_ run it, however, because this is slow.

```{r importing_StormData_read.csv, cache=TRUE, eval=FALSE}
StormData <- read.csv("StormData.csv.bz2", stringsAsFactors = FALSE)
```

The `data.table` package provides a much faster way of getting the data into R, so we
will use that instead:

```{r install_data.table, message=FALSE, warning=FALSE}
options(repos="https://cran.rstudio.com" )
if (!require(data.table)) install.packages("data.table")
library(data.table)
```

Note that `fread()` can not handle the unzipping for us. We will have to use the bunzip2 shell command for that. Luckily `fread()` can handle shell commands and process the output:

```{r importing_StormData_fread, cache=TRUE}
StormData <- fread('bunzip2 -ck StormData.csv.bz2')
```

We've set `cache` to TRUE for the code chunk above to avoid endless importing during debugging, or when running / knitting the document multiple times. Loading the data into R must be cached. Eventhough `fread()` is a lot faster than `read.csv()` (a couple of seconds versus minutes!), it's still a good idea to prevent the code from running repeatedly.




## Results

Is there a section titled "Results" where the main results are presented?

## Review Criteria

1. Has either a (1) valid RPubs URL pointing to a data analysis document for this assignment been submitted; or (2) a complete PDF file presenting the data analysis been uploaded?
2. Is the document written in English?
3. Does the analysis include description and justification for any data transformations?
4. Does the document have a title that briefly summarizes the data analysis?
5. Does the document have a synopsis that describes and summarizes the data analysis in less than 10 sentences?
6. Is there a section titled "Data Processing" that describes how the data were loaded into R and processed for analysis?
7. Is there a section titled "Results" where the main results are presented?
8. Is there at least one figure in the document that contains a plot?
9. Are there at most 3 figures in this document?
10. Does the analysis start from the raw data file (i.e. the original .csv.bz2 file)?
11. Does the analysis address the question of which types of events are most harmful to population health?
12. Does the analysis address the question of which types of events have the greatest economic consequences?
13. Do all the results of the analysis (i.e. figures, tables, numerical summaries) appear to be reproducible?
14. Do the figure(s) have descriptive captions (i.e. there is a description near the figure of what is happening in the figure)?
15. As far as you can determine, does it appear that the work submitted for this project is the work of the student who submitted it?
